A Survey of Exoskeleton Control (Draft 1.1)

### 2015-09-19 paper.tex
more updates

\newcommand{\T}{{\mbox {T}}}

\newcommand{\gc}{{\mbox {GravityCompensation}}}

@@

We do not survey exoskeleton technology for human arms. 
We treat a torso and helmet of the exoskeleton as the payload,
and focus on a lower body exoskeleton to support any payloads that
are on the torso or head.
We do not cover global energy storage such as batteries or liquid fuel.

@@

Hip, knee, ankle: At least 200Nm, ideally >400Nm.

@@

{\it 140 bar(2,050 psi) and 8 l pm in total. Hydraulic cylinders
are designed to be double-acting at the hip pitch joint and
the knee pitch joint. The hip and knee cylinders are designed
for a maximum thrust of 4 kN and 2 kN, respectively. The
four three-way servo valves(M 200, Star-Hydraulic)}~\cite{IEEE07222598}

@@

The dynamics of a revolute exoskeleton are:
\begin{equation}
\mM(\vq) \ddot{\vq} + \mC(\vq,\dot{\vq}) + \mG(\vq) + \mF(\vq,\dot{\vq})
+ \mJ_w^{\tr}(\vq) \vf_w = \vtau + \mJ_o^{\tr}(\vq) \vf_o
\end{equation}
where $\vq$ are the joint angles and position and velocity of the ``root'' of
the exoskeleton, $\mM$ is the inertia matrix, $\mC()$ are the Coriolis and
centripetal forces, $\mG()$ are the gravitational forces, $\mF$ are the
friction forces, $\mJ_w^{\tr}(\vq) \vf_w$ (Jacobian matrix multiplied with the
contact forces)
are the exoskeleton-world
contact forces 
expressed as exoskeleton joint torques, 
$\vtau$ are the exoskeleton
joint torques,
and $\mJ_o^{\tr}(\vq) \vf_o$
are the operator-exoskelton forces expressed
as exoskeleton joint torques,

To understand these equations better, it is useful to have all Jacobians equal
to the identity matrix. This means the operator is directly applying torques
at the exoskeleton joints, and so is the world. In this case:
\begin{equation}
\mM(\vq) \ddot{\vq} + \mC(\vq,\dot{\vq}) + \mG(\vq) + \mF(\vq,\dot{\vq})
+ \vtau_w = \vtau + \vtau_o
\end{equation}

@@

The goal here is to map from desired motion (in this case desired
accelerations) to exoskeleton actuator commands (and desired operator
and world contact forces).
The exoskeleton actuator torque 
attempts to match the torques caused by inertial, Coriolis, centripetal,
gravitational, frictional, and exoskeleton-world contact forces.
It relies on estimated dynamic models and their parameters.
If there are actuator dynamics or other dynamics, we can include those
dynamics as well.
\begin{equation}
\vtau = \vtau_{ID}
\hat{\mM}(\hat{\vq}) \ddot{\vq}_d
+ \hat{\mC}(\hat{\vq},\hat{\dot{\vq}})
+ \hat{\mG}(\hat{\vq})
+ \hat{\mF}(\hat{\vq},\hat{\dot{\vq}})
+ \hat{\mJ}_w^{\tr}(\hat{\vq}) \hat{\vf}_w
\end{equation}
If the models and estimated state (position and velocity) are perfect, we get
perfect cancellation of most of the dynamics:
\begin{equation}
\mM(\vq) (\ddot{\vq} - \ddot{\vq}_d) = \mJ_o^{\tr}(\vq) \vf_o
\end{equation}
If $\ddot{\vq}_d$ is set to zero, the operator directly drives the acceleration
of the exoskeleton with an effective inertia of $(\mM^{-1}(\vq) \mJ_o^{\tr}(\vq))^{-1}$
\begin{equation}
\ddot{\vq} = \mM^{-1}(\vq) \mJ_o^{\tr}(\vq) \vf_o
\end{equation}
$\ddot{\vq}_d$ can be used to drive the exoskeleton along trajectories, and
the operator can directly modify the trajectory.

Simplifying Equation X by assuming the operator directly applies torques at the
exoskeleton joints we get:
\begin{equation}
\mM (\vq) \ddot{\vq} = \vtau_o
\end{equation}
The operator sees the true inertia of the exoskeleton, but no gravitational
loads and no other dynamics (Coriolis, centripetal, or frictional).
Note that we cannot change the apparent inertia of the exoskeleton without
some form of acceleration or force feedback.

@@

To make the exoskeleton appear as a locally linear impedance in joint
coordinates to the operator, we add position and velocity feedback to $\vtau$:
\begin{equation}
\vtau = \vtau_{ID} - \mK_{\vq} ( \hat{\vq} - \vq_d ) - \mK_{\dot{\vq}} \hat{\dot{\vq}}
\end{equation}
so if the dynamic models and state estimation are perfect we get this impedance:
\begin{equation}
\mM \ddot{\vq} - \mK_{\vq} ( \vq - \vq_d ) - \mK_{\dot{\vq}} \dot{\vq} = \vtau_o
\end{equation}
and the operator sees the desirecd impedance.

@@

We can implement force control 
if measurements of the contact forces 
the exoskeleton are available.
The exoskeleton moves as to create as little force as possible 
assuming the actuators are torque sources.
\begin{equation}
\vtau = \vtau_{ID} - \mK_{\vf} \mJ^{\tr} ( \hat{\vf} - \vf_d )
\end{equation}
$\mK_{\vf}$ is a gain matrix, $\mJ$ is the appropriate Jacobian matrix for
force $\vf$
Note that this does not require inverting the Jacobian matrix.
We will see that the position control version does invert the Jacobian matrix.

Applying this to force between the operator and the exoskeleton, and assuming
that the operator directly applies exoskeleton joint torques,
\begin{equation}
\vtau = \vtau_{ID} - \mK_{\vtau_o} ( \hat{\vtau}_o - {\vtau_o}_d )
\end{equation}
So if our dynamic models and state estimation are perfect, we get:
\begin{equation}
\ddot{\vq} = - \mM^{-1} (\vq) \mK_{\vtau_o} ( \hat{\vtau}_o - {\vtau_o}_d )
\end{equation}

One can add force damping terms, which sometimes help:
\begin{equation}
\vtau = \vtau_{ID} - \mK_{\vtau_o} ( \hat{\vtau}_o - {\vtau_o}_d )
- \mK_{\dot{\vtau}_o} \hat{\dot{\vtau}}_o
\end{equation}

Note that impedance control can also be implemented using force control
by setting:
\begin{equation}
{\vtau_o}_d = \mK_{\vq}( \hat{\vq} - \vq_d ) + \mK_{\dot{\vq}} \hat{\dot{\vq}}
\end{equation}
in addition to directly including the stiffness and damping terms in
equation YYY.

@@

The exoskeleton moves as to create as little force as possible between the
operator and the exoskeleton. The actuators in this case may be position sources or
forces sources with a high servo gains,
but ultimately operator-exoskeleton forces are mapped
to exoskeleton velocities or angular velocities.

\begin{equation}
\dot{\vq}_d = \mJ^{-1} (\mK_1 \vf)
\end{equation}
where $\dot{\vq}_d$ are the commanded exoskeleton joint velocities,
$\mJ$ is an appropriate Jacobian matrix, $\mK_1$ is a gain matrix,
$\vf$ is the force vector to be controlled.
[06990981]

Inverting a Jacobian matrix is problematic when the Jacobian matrices are
near singular.
Using a fixed or gain scheduled gain matrix may make more sense.
\begin{equation}
\dot{\vq}_d = \mK_3 \vf
\end{equation}
Simplifying Equation Z by assuming the forces to be controlled
are directly applied to the exoskeleton joints, and thus the Jacobian matrix
is the identity matrix, gives us a similar equation:
\begin{equation}
\dot{\vq}_d = \mK_1 \vtau_o
\end{equation}
and we avoid inverting a Jacobian and problems with singularities.

@@

Sensitivity Amplification Control (SAC) is a dynamic cancellation technique (a.k.a.
inverse dynamics, computed torque, or feedback linearization).

@@

{\bf Lowest level Control:} present interface to actuators and
high bandwidth feedback control, degrees of freedom,
or synergies of either:
\begin{itemize}
\item
Position control.
\item
Pure force or torque control.
\item
Impedance control: specify desired position (for spring), and generate
stiffness, damping, and potentially a modified mass/iinertia/moment of inertia.
Modifying inertia is very difficult.
Or (almost equivalently),
specify a desired force and change the force according to position, velocity,
and acceleration.
\end{itemize} 
One important synergy is coordinated control of the center of mass location and
velocity for balance.
PID or fancier feedback control laws are implemented at this level.
Gravity, friction, actuator dynamics, and other feedforward disturbance
compensation is implemented at this level, as well as any inverse dynamics
or other types of feedforward reference tracking control.
Disturbance observers for model errors and contact/wind forces
are implemented at this level.
Typically this is done by a high rate global servo~\cite{Atlas-robot}
or local limb or joint-level servos~\cite{Sarcos-robot}.

>>>>

\newcommand{\gc}{{\mbox {GravityCompensation}}}
\newcommand{\invdyn}{{\mbox {\tiny ID}}}

@@

We do not survey exoskeleton technology for human arms. 
We treat the torso and helmet of the exoskeleton as the payload,
and focus on a lower body exoskeleton to support any payloads that
are on the torso or head.
We do not cover global energy storage such as batteries or liquid fuel.

@@

Hip, knee, ankle: At least 200Nm, ideally $>$ 400Nm.

@@

{\it 140 bar(2,050 psi) and 8 lpm in total. Hydraulic cylinders
are designed to be double-acting at the hip pitch joint and
the knee pitch joint. The hip and knee cylinders are designed
for a maximum thrust of 4 kN and 2 kN, respectively. The
four three-way servo valves(M 200, Star-Hydraulic)}~\cite{IEEE07222598}

@@

The dynamics of a revolute exoskeleton are:
\begin{equation}
\mM(\vq) \ddot{\vq} + \mC(\vq,\dot{\vq}) + \mG(\vq) + \mF(\vq,\dot{\vq})
+ \mJ_w^{\tr}(\vq) \vf_w = \vtau + \mJ_o^{\tr}(\vq) \vf_o
\end{equation}
where $\vq$ are the joint angles and position and velocity of the ``root'' of
the exoskeleton, $\mM$ is the inertia matrix, $\mC()$ are the Coriolis and
centripetal forces, $\mG()$ are the gravitational forces, $\mF$ are the
friction forces, $\mJ_w^{\tr}(\vq) \vf_w$ (Jacobian matrix multiplied with the
contact forces)
are the exoskeleton-world
contact forces 
expressed as exoskeleton joint torques, 
$\vtau$ are the exoskeleton
joint torques,
and $\mJ_o^{\tr}(\vq) \vf_o$
are the operator-exoskelton forces expressed
as exoskeleton joint torques,
$\vtau$ is augmented with zeros corresponding to the dimensions of $\dot{\vq}$
that correspond to root velocities, since those dimensions are not actuated.

To understand these equations better, it is useful to have all Jacobians equal
to the identity matrix. This means the operator is directly applying torques
at the exoskeleton joints ($\vtau_o$), and so is the world ($\vtau_w$). In this case:
\begin{equation}
\mM(\vq) \ddot{\vq} + \mC(\vq,\dot{\vq}) + \mG(\vq) + \mF(\vq,\dot{\vq})
+ \vtau_w = \vtau + \vtau_o
\end{equation}

@@

The goal here is to map from desired motion (in this case desired
accelerations) to exoskeleton actuator commands (and desired operator
and world contact forces).
The exoskeleton actuator torque 
attempts to match the torques caused by inertial, Coriolis, centripetal,
gravitational, frictional, and exoskeleton-world contact forces.
It relies on estimated dynamic models and their parameters.
If there are actuator dynamics or other dynamics, we can include those
dynamics as well.
\begin{equation}
\vtau = \vtau_{\invdyn} 
= \hat{\mM}(\hat{\vq}) \ddot{\vq}_d
+ \hat{\mC}(\hat{\vq},\hat{\dot{\vq}})
+ \hat{\mG}(\hat{\vq})
+ \hat{\mF}(\hat{\vq},\hat{\dot{\vq}})
+ \hat{\mJ}_w^{\tr}(\hat{\vq}) \hat{\vf}_w
\end{equation}
If the models and estimated state (position and velocity) are perfect, we get
perfect cancellation of most of the dynamics:
\begin{equation}
\mM(\vq) (\ddot{\vq} - \ddot{\vq}_d) = \mJ_o^{\tr}(\vq) \vf_o
\end{equation}
If $\ddot{\vq}_d$ is set to zero, the operator directly drives the acceleration
of the exoskeleton with an effective inertia of $(\mM^{-1}(\vq) \mJ_o^{\tr}(\vq))^{-1}$
\begin{equation}
\ddot{\vq} = \mM^{-1}(\vq) \mJ_o^{\tr}(\vq) \vf_o
\label{eq:cancel}
\end{equation}
$\ddot{\vq}_d$ can be used to drive the exoskeleton along trajectories, and
the operator can directly modify the trajectory.

Simplifying equation~\ref{eq:cancel} by 
assuming the operator directly applies torques at the
exoskeleton joints we get:
\begin{equation}
\mM (\vq) \ddot{\vq} = \vtau_o
\end{equation}
The operator sees the true inertia of the exoskeleton, but no gravitational
loads and no other dynamics (Coriolis, centripetal, or frictional).

Note that we cannot change the apparent inertia of the exoskeleton without
some form of acceleration or force feedback.
With acceleration feedback we can make the exoskeleton have a different inertia.
Setting $\ddot{\vq}_d$ to zero and adding acceleration feedback:
\begin{equation}
\vtau = \vtau_{\invdyn} - (\mM_d(\hat{\vq}) - \hat{\mM}(\hat{\vq})) \hat{\ddot{\vq}}
\label{eq:change-inertia1}
\end{equation}
sets the apparent inertia to $\mM_d$ if the dynamic models are perfect, and
the state estimation produces not only perfect positions and velocities but
also perfect acceleration estimates. In the case 
where the operator directly applies torques at the
exoskeleton joints we get:
\begin{equation}
\mM_d (\vq) \ddot{\vq} = \vtau_o
\end{equation}
In robotics this is usually considered a bad idea if $\mM_d$ is smaller (a complex
issue for a matrix) than $\mM$ at any $\vq$. This type of control is vulnerable
to unmodeled dynamics. However, in an exoskeleton, since we have not tried to
cancel the human's dynamics, the human stabilizes a possible unstable exoskeleton
(if the straps are tight enough).

With force feedback we can also make the exoskeleton have a different inertia.
Setting $\ddot{\vq}_d$ to zero and adding force feedback:
\begin{equation}
\vtau = \vtau_{\invdyn} + \mK \hat{\vf}_o
\end{equation}
In the case where the force measurements are perfect and 
the operator directly applies torques at the
exoskeleton joints we get:
\begin{equation}
(\mK + \mI)^{-1} \mM (\vq) \ddot{\vq} = \vtau_o
\end{equation}
By setting $\mK = \mM^{-1}_d (\vq) \mM (\vq) - \mI$ we can achieve
\begin{equation}
\mM_d (\vq) \ddot{\vq} = \vtau_o
\end{equation}

@@

To make the exoskeleton appear as a locally linear impedance in joint
coordinates to the operator, we add position and velocity feedback to $\vtau$:
\begin{equation}
\vtau = \vtau_{\invdyn} - \mK_{\vq} ( \hat{\vq} - \vq_d ) - \mK_{\dot{\vq}} \hat{\dot{\vq}}
\label{eq:impedance}
\end{equation}
so if the dynamic models and state estimation are perfect we get this impedance:
\begin{equation}
\mM \ddot{\vq} - \mK_{\vq} ( \vq - \vq_d ) - \mK_{\dot{\vq}} \dot{\vq} = \vtau_o
\end{equation}
and the operator sees the desirecd impedance.

@@

We can implement force control 
if measurements of the contact forces with
the exoskeleton are available.
The exoskeleton moves as to create as little force as possible 
assuming the actuators are torque sources.
\begin{equation}
\vtau = \vtau_{\invdyn} - \mK_{\vf} \mJ^{\tr} ( \hat{\vf} - \vf_d )
\end{equation}
$\mK_{\vf}$ is a gain matrix, $\mJ$ is the appropriate Jacobian matrix for
force $\vf$
Note that this does not require inverting the Jacobian matrix.
We will see that the position control version does invert the Jacobian matrix.

Applying this to force between the operator and the exoskeleton, and assuming
that the operator directly applies exoskeleton joint torques,
\begin{equation}
\vtau = \vtau_{\invdyn} - \mK_{\vtau_o} ( \hat{\vtau}_o - {\vtau_o}_d )
\end{equation}
So if our dynamic models and state estimation are perfect, we get:
\begin{equation}
\ddot{\vq} = - \mM^{-1} (\vq) \mK_{\vtau_o} ( \hat{\vtau}_o - {\vtau_o}_d )
\end{equation}

One can add force damping terms, which sometimes help:
\begin{equation}
\vtau = \vtau_{\invdyn} - \mK_{\vtau_o} ( \hat{\vtau}_o - {\vtau_o}_d )
- \mK_{\dot{\vtau}_o} \hat{\dot{\vtau}}_o
\end{equation}

Note that impedance control can also be implemented using force control
by setting:
\begin{equation}
{\vtau_o}_d = \mK_{\vq}( \hat{\vq} - \vq_d ) + \mK_{\dot{\vq}} \hat{\dot{\vq}}
\end{equation}
in addition to directly including the stiffness and damping terms in
equation~\ref{eq:impedance}.

@@

The exoskeleton moves as to create as little force as possible between the
operator and the exoskeleton. The actuators in this case may be position sources or
force sources with high servo gains,
but ultimately operator-exoskeleton forces are mapped
to exoskeleton velocities or angular velocities.

\begin{equation}
\dot{\vq}_d = \mJ^{-1} (\mK_1 \vf)
\label{eq:gootwcps}
\end{equation}
where $\dot{\vq}_d$ are the commanded exoskeleton joint velocities,
$\mJ$ is an appropriate Jacobian matrix, $\mK_1$ is a gain matrix,
and $\vf$ is the force vector to be controlled~\cite{IEEE06990981}.

Inverting a Jacobian matrix is problematic when the matrix is
nearly singular.
Using a fixed or gain scheduled gain matrix may make more sense.
\begin{equation}
\dot{\vq}_d = \mK_3 \vf
\end{equation}
Simplifying equation~\ref{eq:gootwcps}
by assuming the forces to be controlled
are directly applied to the exoskeleton joints, and thus the Jacobian matrix
is the identity matrix, gives us a similar equation:
\begin{equation}
\dot{\vq}_d = \mK_1 \vtau_o
\end{equation}
and we avoid inverting a Jacobian and problems with singularities.

@@

Sensitivity Amplification Control (SAC) is a dynamic cancellation technique (a.k.a.
inverse dynamics, computed torque, or feedback linearization) that also tries
to adjust the apparent inertia of the exoskeleton (equation~\ref{eq:change-inertia1})

@@

{\bf Lowest level Control:} present interface to actuators and
high bandwidth feedback control, degrees of freedom,
or synergies of either:
\begin{itemize}
\item
Position control.
\item
Pure force or torque control.
\item
Impedance control: specify desired position (for spring), and generate
stiffness, damping, and potentially a modified inertia or moment of inertia.
Modifying inertia is more dangerous than providing active stiffness
or damping.
Or (almost equivalently),
specify a desired force and change the force according to position, velocity,
and acceleration.
\end{itemize} 
One important synergy is coordinated control of the center of mass location and
velocity for balance.
PID or fancier feedback control laws are implemented at this level.
Gravity, friction, actuator dynamics, and other feedforward disturbance
compensation is implemented at this level, as well as any inverse dynamics
or other types of feedforward reference tracking control.
Disturbance observers for model errors and contact/wind forces
are implemented at this level.
Typically this is done by a high sampling rate global servo~\cite{Atlas-robot}
or local limb or joint-level high sampling rate servos~\cite{Sarcos-robot}.
